An agent-based model for city networks based on interactions between firms

### 2020-03-05 main.tex
Update on Overleaf.

\begin{abstract}
    This paper introduces a generative model for urban networks defined by interactions between firms. The aim is to test several geographical properties emerging in such a network as metropolisation, internationalisation and regionalisation and test the effect of specialisation on these phenomena. The growth of the network is simulated by adding randomly new links defined by ownership between firms, with probabilities of occurring depending on the economic size of urban areas at origin and destination, industrial sector proximity between firms, the strength of links from the past as well as the geographical and socio-cultural distance. The simulation on a synthetic system of cities unveil phase transitions when changing interaction distance, and non-trivial patterns in the model behavior. Calibration of the model on real data on company ownership linkages for Europe in 2018 is also undertaken. Future work will consist in observing thanks to this model the impact of economic crisis on such networks.

@@

The intensification of globalisation of the economy since the late 20th century resulted in the rise of the network society \cite{castells2000networksociety} and the emergence of a world-city system \cite{taylor2001specification}, characterized by highly interconnected urban centres at the global level \cite{sassen1991}. These interactions can be of different nature and determined by economic, socio-cultural, or geopolitical proximity implying spatial and non spatial patterns \cite{martinus2018global}. Interactions of economic nature between transnational firms, are together with trade among the most representative of the current economic trends \cite{taylor2001specification}. Understanding the drivers of growth and geographical properties of such a urban network, can be used to foster innovation in specific cities and to shape public policies for local industrial clusters \cite{turkina2016structure}. Straightening these city-interactions can be crucial for revitalising certain regions \cite{Clarke2018}, by developing strategic industries for global insertion of regions \cite{dawley2019creating}, cities \cite{gluckler2016relational} and countries \cite{martinus2019brokerage}. The functioning of such a city-network and the position of cities within these networks \cite{gluckler2016relational} can also permit to infer the consequences of future exogenous shocks such as an economic crisis, a single market collapse or a post-Brexit scenario for the European Union.

The use of network models to understand processes underlying the emergence and dynamics of such urban networks has already been widely tackled in the literature. \cite{dai2016simulating} propose a generative urban network model at the regional scale, combining geographical factors with network topological properties. Air transportation networks can also be understood as urban networks, and \cite{zanin2013modelling} reviews several models for such networks. Trade network have been extensively studied from complex networks perspective, but are generally driven at the country-level \cite{garlaschelli2005structure} because of lack of city-level data. Input-output models, mostly used in Regional science to represent flows between geographical areas, are considered in some cases as a type of urban generative model \cite{jin1993generation}. In this stream of research, spatial interaction models \cite{dennett2013multilevel} can form the basis of urban network models for abstract flows \cite{dai2016generative} or physical infrastructure \cite{raimbault2020indirect}.

@@

Companies from the AMADEUS database are georeferenced using the Geonames database (using postcode or address depending on the information available). We then join this data with the GHS-FUA dataset for functional urban areas \cite{Florczyk2019ghs}, which are the relevant objects for applying our model. Starting from a firm-level dataset of 1,562,862 nodes and  1,866,936 links, we obtain a directed network of 573 urban areas and 9323 ownership links. Weight of links are obtained by computing the owned share of turnover at destination, i.e. $w_{ij} = \sum_{k \in i,l \in j} p_{kl} \cdot T_l$ where $p_{kl}$ of the ownership share of company $l$ by company $k$ and $T_l$ the turnover of company $l$. Node attributes include an economic size of urban areas, that we compute as the sum of turnovers of companies within the area. This quantity is highly correlated with area-level GDP included in the GHS database (Spearman correlation $\rho = 0.71$) and also with population ($\rho = 0.60$), so we expect to capture size effects while keeping the consistence of a single main data source. An other attribute of nodes is the industrial sector composition, expressed as a proportion of turnover in the area associated to a given industrial sector. Detailed sectors of firms are available up to the fourth digit for the NACE classification in the raw data. We consider the first level classification (21 categories) and compute sector proportions accordingly. This allows us to define an industrial proximity matrix between urban areas, taking a cosine similarity between vectors as $s_{ij} = \sum_{k=1}^{K} S_{ik}\cdot S_{jk}$ where $S_{ik}$ are sector proportions.

@@

Regarding the community structure of the network, different modularity maximization algorithms (run on the undirected corresponding network) give a same undirected modularity of 0.38, and a directed modularity \cite{nicosia2009extending} of 0.34 for the greedy algorithm of \cite{clauset2004finding} and 0.36 for the Louvain algorithm \cite{blondel2008fast}, corresponding to 36 communities (resp. 15). In comparison, with 29 countries, the classification of urban areas by country has a modularity of 0.32. This means that the structure of flows has a certain regional integration. Indeed, a null model assigning random labels with 30 communities, bootstrap 1000 times, yields an average modularity of $0.049 \pm 0.002$, confirming the statistical significance of these communities.

To have a first insight into factors playing a role in the formation of links, we proceed to a statistical analysis which is similar to unconstrained spatial interaction models \cite{wilson1975some}. We test five different statistical models, including progressively the processes candidate to be included in the network generative model. The most general statistical model is
\[
\log(w_{ij}) = \log(d_{ij}) + \log(T_i) + \log(T_j) + \log(s_{ij}) + \alpha_{c_i,c_j} + \varepsilon_{ij}
\]
where $\alpha_{c_i,c_j}$ is a fixed effect term for the couple of countries $c_i,c_j$. We use a multiplicative form by taking logarithms as it is usually done for spatial interaction models. Furthermore, origin and/or destination constrained models are not relevant in our case as links include not only turnover but also a share of ownership. Estimation results are given in Table~\ref{tab:reg}. We also fit a Poisson generalized linear model with a logarithmic link on the rounded weights \cite{flowerdew1988fitting}. We find that distance alone (model 1) has a significant negative effect but with an explained variance close to zero. The second model adds fixed effects by pair of countries, for which around a third are significant, and with a better $R^2$ of 0.17. Adding origin and destination economic size (turnovers $T_i$ and $T$, model 3) slightly improve the explanative power, but the best model among the linear ones is the model 4 with all factors and fixed effects, both in terms of explained variance ($R^2 = 0.31$) and Akaike information criterion (what means that there is no overfitting). The Poisson model with all variables is much better regarding pseudo-$R^2$ and provides much more significant estimates (all fixed effects significant and much smaller standard deviations). Across all models, we find consistent qualitative stylized facts: (i) distance has a negative influence; (ii) economic sizes have a positive influence, with the size of the origin being more important (what is consistent with ownership links, strongly driven by decision at the headquarters); (iii) industrial proximity has a positive influence, i.e. similarity is more important than complementarity for this dimension; (iv) an important part of country fixed-effects are significant; and (v) the order of magnitude of the influence of each factor is roughly the same, in the sense that no process totally dominates the others. This suggests the relevance of including all these different dimensions in the generative network model.

>>>>

\begin{abstract}
    This paper introduces a generative model for urban networks defined by interactions between firms to test several geographical properties emerging in such a network as metropolisation or internationalisation. The growth of the network is simulated by adding randomly new links defined by ownership between firms, with probabilities of occurring depending on the economic size of urban areas at origin and destination, industrial sector proximity between firms, the strength of links from the past as well as the geographical and socio-cultural distance. The simulation on a synthetic system of cities unveil phase transitions when changing interaction distance, and non-trivial patterns in the model behavior. Calibration of the model on real data on company ownership linkages for Europe in 2018 is also undertaken. Future work will consist in observing thanks to this model the impact of economic crisis on such networks.

@@

The intensification of globalisation of the economy since the late 20th century resulted in the rise of the network society \cite{castells2000networksociety} and the emergence of a world-city system \cite{taylor2001specification}, characterized by highly interconnected urban centres at the global level \cite{sassen1991}. These interactions can be of different nature and determined by economic, socio-cultural, or geopolitical proximity implying spatial and non spatial patterns \cite{martinus2018global}. Interactions of economic nature between transnational firms, are together with trade among the most representative of the current economic trends \cite{taylor2001specification}. Understanding the drivers of growth and geographical properties of such a urban network, can be used to foster innovation in specific cities and to shape public policies for local industrial clusters \cite{turkina2016structure}. Straightening these city-interactions can be crucial for revitalising certain geographical areas \cite{Clarke2018}, by developing strategic industries for global insertion of regions \cite{dawley2019creating}, cities \cite{gluckler2016relational} and countries \cite{martinus2019brokerage}. The functioning of such a city-network and the position of cities within these networks \cite{gluckler2016relational} can also permit to infer the consequences of future exogenous shocks such as an economic crisis, a single market collapse or a post-Brexit scenario for the European Union.

The use of network models to understand processes underlying the emergence and dynamics of such urban networks has already been widely tackled in the literature. \cite{dai2016simulating} propose a generative urban network model at the regional scale, combining geographical factors with network topological properties. Air transportation networks can also be understood as urban networks, and \cite{zanin2013modelling} reviews several models for such networks. Trade network have been extensively studied from complex networks perspective, but are generally driven at the country-level \cite{garlaschelli2005structure} because of lack of city-level data. Input-output models, mostly used in regional science to represent flows between geographical areas, are considered in some cases as a type of urban generative model \cite{jin1993generation}. In this stream of research, spatial interaction models \cite{dennett2013multilevel} can form the basis of urban network models for abstract flows \cite{dai2016generative} or physical infrastructure \cite{raimbault2020indirect}.

@@

Companies from the AMADEUS database are georeferenced using the Geonames database (using postcode or address depending on the information available). We then join this data with the GHS-FUA dataset for Functional Urban Areas \cite{Florczyk2019ghs}, which are the most relevant objects for applying our model. Starting from a firm-level dataset of 1,562,862 nodes and  1,866,936 links, we obtain a directed network of 573 urban areas and 9323 ownership links. Weight of links are obtained by computing the owned share of turnover at destination, i.e. $w_{ij} = \sum_{k \in i,l \in j} p_{kl} \cdot T_l$ where $p_{kl}$ of the ownership share of company $l$ by company $k$ and $T_l$ the turnover of company $l$. Node attributes include an economic size of urban areas, that we compute as the sum of turnovers of companies within the area. This quantity is highly correlated with area-level GDP included in the GHS database (Spearman correlation $\rho = 0.71$) and also with population ($\rho = 0.60$), so we expect to capture size effects while keeping the consistence of a single main data source. An other attribute of nodes is the industrial sector composition, expressed as a proportion of turnover in the area associated to a given industrial sector. Detailed sectors of firms are available up to the four digits of the NACE classification in the raw data. We consider the first level classification (21 categories) and compute sector proportions accordingly. This allows us to define an industrial proximity matrix between urban areas, taking a cosine similarity between vectors as $s_{ij} = \sum_{k=1}^{K} S_{ik}\cdot S_{jk}$ where $S_{ik}$ are sector proportions.

@@

Regarding the community structure of the network, different modularity maximization algorithms (run on the undirected corresponding network) give a same undirected modularity of 0.38, and a directed modularity \cite{nicosia2009extending} of 0.34 for the greedy algorithm of \cite{clauset2004finding} and 0.36 for the Louvain algorithm \cite{blondel2008fast}, corresponding to 36 communities (resp. 15). In comparison, with 29 countries, the classification of urban areas by country has a modularity of 0.32. This means that the structure of flows present certain regional patterns. Indeed, a null model assigning random labels with 30 communities, bootstrap 1000 times, yields an average modularity of $0.049 \pm 0.002$, confirming the statistical significance of these communities.

To have a first insight into factors playing a role in the formation of links, we proceed to a statistical analysis similar to unconstrained spatial interaction models \cite{wilson1975some}. We test five different statistical models, including progressively the processes candidate to be included in the network generative model. The most general statistical model is
\[
\log(w_{ij}) = \log(d_{ij}) + \log(T_i) + \log(T_j) + \log(s_{ij}) + \alpha_{c_i,c_j} + \varepsilon_{ij}
\]
where $\alpha_{c_i,c_j}$ is a fixed effect term for the couple of countries $c_i,c_j$. We use a multiplicative form by taking logarithms as it is usually done for spatial interaction models. Furthermore, origin and/or destination constrained models are not relevant in our case as links include not only turnover but also a share of ownership. Estimation results are given in Table~\ref{tab:reg}. We also fit a Poisson generalized linear model with a logarithmic link on the rounded weights \cite{flowerdew1988fitting}. We find that distance alone (model 1) has a significant negative effect but with an explained variance close to zero. The second model adds fixed effects by pair of countries, for which around a third are significant, and with a better $R^2$ of 0.17. Adding origin and destination economic size (turnovers $T_i$ and $T$, model 3) slightly improve the explanative power, but the best model among the linear ones is the model 4 with all factors and fixed effects, both in terms of explained variance ($R^2 = 0.31$) and Akaike information criterion. The la(what means that there is no overfitting). The Poisson model with all variables is much better regarding pseudo-$R^2$ and provides much more significant estimates (all fixed effects significant and much smaller standard deviations). Across all models, we find consistent qualitative stylized facts: (i) distance has a negative influence; (ii) economic sizes have a positive influence, with the size of the origin being more important (what is consistent with ownership links, strongly driven by decision at the headquarters); (iii) industrial proximity has a positive influence, i.e. similarity is more important than complementarity for this dimension; (iv) an important part of country fixed-effects are significant; and (v) the order of magnitude of the influence of each factor is roughly the same, in the sense that no process totally dominates the others. This suggests the relevance of including all these different dimensions in the generative network model.
